---
title: "STATS 531 Midterm Project"
author: "Mukai Wang 98830336"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(lubridate)
library(dplyr)
library(ggplot2)
library(reshape2)
```


# Introduction
Keeping track of the price of industrial scale food is an important measure to regulate the food market. [The Mandatory Price Reporting Act of 2010](https://www.congress.gov/111/plaws/publ239/PLAW-111publ239.pdf) requires the USDA to release dairy product sales information on or before Wednesday at 3:00 pm EST (unless affected by a Federal Holiday). This weekly reported price information offers insight into time trend of several dairy products in the past few years.

The data is collected from [Kaggle](https://www.kaggle.com/sohier/weekly-dairy-product-prices) and belongs to [USDA](https://mpr.datamart.ams.usda.gov/menu.do?path=\Products\Dairy). The data contain prices (dollars per pound) of five different kinds of dairy products from March 2012 to December 2017. In this project, I am going to focus on the fluctuation of butter price. 

# Data Cleaning and Exploration

The data cleaning addresses several issues with the raw data,

* There are duplicate entries in the dataset that I need to remove.
* There are several entries of prices taken on the same day. My decision is to aggregate these prices and take the average.
* Some adjacent time points don't differ by 7 days. My approach is to only retain those time points whose date differences from the very first day in the dataset (2012/03/03) are multiples of 7.


```{r, echo=FALSE}
butter = read.csv("https://raw.githubusercontent.com/skybullbobby/Weekly-Dairy-Product-Time-Series-Analysis/master/weekly-dairy-product-prices/Butter.csv")
```

```{r, echo=FALSE}
data_process = function(rawdata){
  # fix the date time formats of three columns
  rawdata$Week.Ending.Date = as.Date(rawdata$Week.Ending.Date, format='%m/%d/%Y')
  rawdata$Report.Date = as.Date(rawdata$Report.Date, format='%m/%d/%Y')
  rawdata$Date = as.Date(rawdata$Date, format='%m/%d')
  year(rawdata$Date) = year(rawdata$Week.Ending.Date)
  # only need the date and weighted prices
  subsetdata = rawdata[,c("Date","Weighted.Prices")]
  # remove duplicate values
  removeduplicates = unique(subsetdata)
  rankeddata = removeduplicates[order(removeduplicates$Date),]
  # there are some values collected at the same date, so I aggregate them by averaging them
  summarized_rankeddata = aggregate(rankeddata$Weighted.Prices,list(rankeddata$Date), mean)
  colnames(summarized_rankeddata) = c("Date", "Price")
  return(summarized_rankeddata)
}
processed_butter = data_process(butter)
```


```{r, echo=FALSE}
dateranges = seq(processed_butter$Date[1], by="week", length = nrow(processed_butter))
butterprice = processed_butter %>% filter(Date %in% dateranges)
```

After my data processing, I have 286 data points from 2012/03/03 to 2017/08/19. I plot the price trend below.

```{r, echo=FALSE, fig.width=10, fig.height=7}
ggplot(butterprice, aes(x=Date, y=Price))+geom_line()+ggtitle("Butter Price (dollar per pound)")
```

Based on this plot, I can confidently say that there is definitely strong trend in the data (not simply white noise). There are large short term fluctuations together with a seemingly increasing trend on a global scale.

# Methodology

Based on the plot of the raw prices above, I decide on the analysis scheme as follows:

1. Decompose the price trend into a long term trend, a short term trend and noise using loess smoothing with two different spans.
2. For the short term trend, I plan to carry out spectrum analysis and identify potential significant frequencies. 
3. For the long term trend, I plan to use ARMA models and their derivatives(SARMA, regression with ARMA errors) to explain the time series trend parametrically.



# Results

## Spectral Analysis for Short Term Trend

```{r, echo=FALSE, fig.width=6, fig.height=9}
date = seq(1,length(butterprice$Price))
raw =ts(butterprice$Price, start=decimal_date(ymd("2012-03-03")), frequency = 365.25/7)
long_term <- ts(loess(butterprice$Price~date,span=0.5)$fitted,
                 start=decimal_date(ymd("2012-03-03")), frequency = 365.25/7)
noise <- ts(butterprice$Price - loess(butterprice$Price~date,span=0.1)$fitted,
                start=decimal_date(ymd("2012-03-03")), frequency = 365.25/7)
short_term <- raw - long_term - noise

plot(ts.union(raw, long_term, short_term, noise), main="Butter price trend")
```


```{r, echo=FALSE}
bspectrum = spectrum(short_term, spans=c(3,5,3), main="Smoothed Spectrum for Short Term Butter Prices Fluctuation")
abline(h=0.5e-2,lty="dashed",col="red")
```


## ARMA Models for Long Term Trend

```{r}
long_term = as.vector(long_term)
slr = lm(long_term~date)
summary(slr)
```





```{r}
slrresiduals = slr$residuals
plot(slrresiduals, main="Residual Scatter Plot")
```


```{r}
acf(slr$residuals, lag.max=100)
```

```{r}
aic_table <- function(data ,P, Q){
  table <- matrix(NA,(P+1),(Q+1))
  for(p in 0:P){
    for(q in 0:Q){
      table[p+1, q+1] <- arima(data,order=c(p,1,q))$aic
    }
  }
  dimnames(table) <- list(paste("AR",0:P, sep=""),paste("MA",0:Q,sep=""))
  table
}

# search for optimal parameter
butter_long_aic <- aic_table(slrresiduals, 4, 5)


```


```{r}
model <- arima(slrresiduals, order=c(2,1,1))
acf(model$residuals, lag.max=52)
```


```{r}
model2 <- arima(slrresiduals, order=c(2,1,1),
  seasonal=list(order=c(2,0,0), period=18))
```

```{r}
acf(model2$residuals, lag.max=52)
model2
```

# Discussion

# References

